Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-05-12 analysis/paper/paper.Rmd
Further interpretation. Identified logistic base camps through the curation index?

This study employs multivariate exploratory statistics to analyse the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway. This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space. However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g. @dibble2017], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites. The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials. Finally, analytical and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined. Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern.

However, even though each individual assemblage can have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages that is based on the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the whole assemblage behaviour index, which has not seen application in this context before.

@@

The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway \ref\@(fig:dist)A]. The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b]. The sites included in the analysis consists of all Mesolithic sites excavated in conjunction with the projects that have assemblages holding more than 100 artefacts. The institution responsible for these excavations was the Museum of Cultural History in Oslo. This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations. Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the museum. As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [e.g. @clark2006; @dibble2017]. In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014, see also \ref\@(fig:dist)B]. Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages. While this is difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis.

```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis. Radiocarbon probability density functions are given as the posterior sum. Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators. These follow the original reports.", fig.height = 11, fig.fullwidth = TRUE}
# Create a bounding box around site points to limit the overview map
sitbbox <- st_bbox(site_pts)
sitbbox[1:2] <- sitbbox[1:2] - 600000
sitbbox[3:4] <- sitbbox[3:4] + 600000
boundingpoly <- st_as_sf(st_as_sfc(sitbbox))

@@

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [@glørstad2010]. Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @sørensen1979; @romundset2018]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic [@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.

@@

Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [e.g. @helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or been concerned with technological processes associated with certain sub-categories of the site inventories, such as blades or axes [e.g. @damlien2016; @solheim2020]. Some studies have involved in-depth analyses of handful of sites in terms of intrasite spatial organisation [@nærøy2018], and technological organisation [@eigeland2015; @mansrud2016;]. There are also some more comprehensive evaluations of a larger number of entire lithic assemblages to get at potential site types and associated mobility patterns [@breivik2016; @viken2018c; @breivik2020]. However, the narratively driven analyses of these data makes the weighting of different variables unclear, and these studies draw their conclusions based on an arguably over-constraining predefined trichotomy of site types. Each site could have had complex life histories and been associated with a wide range of tasks that in sum make up the assemblages. Finally, only a single study concerned with eight Middle-Mesolithic assemblages --- assemblages that are also treated here --- has employed a multivariate quantitative framework to help structure such a treatment [@solheim2013]. In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods.

In a series of case studies from Northern Norway, @bølviken1982 were among a group of scholars often attributed the popularisaton of correspondence analysis (CA) in wider archaeologcal circles [@baxter1994]. In southern Norway, however, the use of multivariate statistics in Mesolithic research is to my knowledge limited to @solheim2013a use of CA to analyse the assemblage data from eight sites excavated in connection with a development-led project. As these sites, which are included in this study, have relatively contemporaneous dates, the Solheims analysis was mainly aimed at identifying potential variation in the

There have been a few studies that have conducted comprehensive analyses of artefact inventories and site features to get at variation in Mesolithic site types [@viken2018; @breivik2020]. As Binford originally clarified, a point that has also been retiterated by several authors since; these categories are best understood as extremes on a continuous scale.

## Methods

The first part of the analysis conducted here involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods when the entirety of the assemblages are taken into account. This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination [e.g. @barton1991; @dibble1995]. Consequently, several artefact categories have here been collapased for the CA. This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading. These have all been combined into the single category "small flint tools". (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption is therefore that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While ultimately intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.

However, the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for does have major implications that the above-outlined analysis does not take properly into account, and renders it difficult to align any identified patterns with specific behavioural dimensions [@dibble2017]. For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life. While use-wear analysis could potentially offer a way to identify what artefacts were used for towards the end of their use-life, these kinds of analyses are extremely time-consuming and are therefore typically only conducted on a smaller number of artefacts that have already been selected for analysis based on their shape [e.g. @solheim2018]. Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017]. As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with these inferential limitations in mind [see @clark2017, and below]. The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1979]. An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of tool replacement. Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@binford1979]. Although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is therefore typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool or extending its use-life by the repeated rejuvination and modification of edges [e.g. @bamforth1986; @dibble1995; @shott2005]. The continuum between curated and expedient technology has in turn been related to land-use and mobility strategies, as well as raw-material quality and availability [e.g. @parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017].

While the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed [e.g. @bamforth1986; @shott1996; @surovell2009,9--13], following not least from the ambiguous definition first put forward by @binford1979. Some concrete operationalisation of the terms have, however, been suggested. Through a series of studies Barton and collegues [e.g. -@barton1999; @barton2014; @clark2017] have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent results across a wide range of chronological and cultural context, and is measure taken to reflect degree of curation associated with the assemblages. which in turn is mainly to follow from variation in mobility patterns associated with the sites [@barton2014]. Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is in this model assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics and a low relative frequency of retouched artefacts. A further empirical expectation of an expendient tehcnological organisation is as a consequence also a higher number of cores and unretouched flakes and blades [@clark2017].

This is the main measure employed here. Some additional variables suggested by @Bicho2020 to be related to mobility patterns are subsequently also explored using principal components analysis (PCA), following the continuous operationalisations of these measures [cf. @baxter1994, 100].

A note should also be made on the fact that a couple of variables that are sometimes invoked for the classification of sites in terms of associated mobility pattern are omitted here. This pertains to number of site features such as fireplaces, cooking pits and dwelling structures. Feature count has been omitted as taphonomic loss is likely to lead to a chronological bias in their preservation. Similarly, the number of activity areas, effectively number artefact clusters, however defined, has also been disregarded. This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied. This pertains for example to the impact of bioturbation in the form of three-throws, which can can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jørgensen2017; @darmark2018e].

Within this framework, amount of retouch has been seen as a empirical indicator of degree of curation. While various methods for measuring the curation associated with individual tools have been forwarded, the approach taken here focuses on the assemblage considered as a whole.

@@

Figure \@ref(fig:cor) displays the CA using the lithic count data. While no clear-cut clusters can be readily delineated, the general impression from the plots is that a chronological dimension is associated with the patterning in the data. This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data. The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension. Although fewer in number, the sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.

@@

lithics2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
          -askeladden_id) %>% 
  # The reason for not using .keep below is for the option to 
  #retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         blanks_freq = sum(blade, microblade, fragment, flake, flake_other, 
                     fragment_other, blade_other,  microblade_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         nonflint_freq = nonflint_count / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
         -chip_other, -core_other, -core_fragment_other) %>% 
  column_to_rownames("site")

# Repeat but for flint only
flint2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project, 
         -reported_start_bc, -reported_end_bc, -askeladden_id) %>% 
  select(!ends_with("other")) %>% 
  mutate(chip_freq =  sum(chip) / artefact_count,
         core_freq = sum(core) / artefact_count,
         blanks_freq = sum(blade, microblade, fragment, flake) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch) / artefact_count,
         flint_dens = flint_count / excavated_m3) %>%
  select(-blanks_flint, -secondary_flint, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -grinding_slab_stone, -core, 
         -core_fragment, -chip, dating_method) %>% 
  column_to_rownames("site")

@@

```{r cor-plot, fig.cap = "Correlation plot showing transformation of skewed variables for the PCA, and the linear correlation between these", fig.height = 9.5, fig.width = 10}
# Transform the skewed variables - sqrt() is used where there are zero values,
# and seems to provide an adequate result.
lithicstr <- lithics2 %>%
  # Transform skewed variables and reorder
  mutate("Mean date cal. BCE*-1" = mean_date * -1,
          "log10(lithic_dens)" = log10(lithic_dens),
         "log10(secondary_freq)" = log10(secondary_freq),
         "core_freq^(1/3)" = core_freq^(1/3),
         "nonflint_freq^(1/3)" = nonflint_freq^(1/3),  .keep = "unused") %>% 
  relocate("Mean date cal. BCE*-1",
    "log10(lithic_dens)", "blanks_freq", "log10(secondary_freq)",
    "chip_freq", "core_freq^(1/3)" 
  )

>>>>

This study employs multivariate exploratory statistics to analyse the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway. This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time. However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g. @dibble2017; @rezek2020], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites. The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials. Finally, analytical and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined. Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern. In an attempt to limit the influence of confounding effects, the material chosen for analysis has a relatively constrained geographical distribution, and stems from relatively recent investigations that have employed comparable methods for excavation and classification.

Even though each individual assemblage can have been impacted by an virtual infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports, which consequently assumes that these are behaviourally meaningful. The second involves an analysis of the assemblages in light of the whole assemblage behaviour index, which has not seen application in this context before.

@@

The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway \ref\@(fig:dist)A]. The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b]. The sites included in the analysis consists of all Mesolithic sites excavated in conjunction with the projects that have assemblages holding more than 100 artefacts. The institution responsible for these excavations was the Museum of Cultural History in Oslo. This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations. Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the museum. As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [e.g. @clark2006; @dibble2017]. In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014, see also \ref\@(fig:dist)B]. Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages. While this is difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects.

```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis. Radiocarbon probability density functions are given as the posterior sum. Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators. These follow the original reports.", fig.height = 10, fig.width = 12}
# Create a bounding box around site points to limit the overview map
sitbbox <- st_bbox(site_pts)
sitbbox[1:2] <- sitbbox[1:2] - 600000
sitbbox[3:4] <- sitbbox[3:4] + 600000
boundingpoly <- st_as_sf(st_as_sfc(sitbbox))

@@

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [@glørstad2010]. Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @sørensen1979; @romundset2018]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of relative sea-level change allows for a relatively good chronological control of when these accumulation events occurred. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic [@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.

@@

Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [e.g. @helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or been concerned with technological processes associated with certain sub-categories of the site inventories, such as blades or axes [e.g. @damlien2016; @berg-hansen2017; @solheim2020]. Some studies have involved in-depth analyses of handful of sites in terms of intrasite spatial organisation [@nærøy2018], and technological organisation [@eigeland2015; @mansrud2016;]. There are also some more comprehensive evaluations of a larger number of sites with associated lithic assemblages to get at potential site types and related mobility patterns [@breivik2016; @viken2018c; @breivik2020]. However, the narratively driven analyses of these data makes the weighting of different variables unclear, and these studies draw their conclusions based on an arguably over-constraining predefined trichotomy of different site types. Finally, only a single study concerned with eight Middle-Mesolithic sites has employed a multivariate quantitative framework to help structure such a treatment [@solheim2013a]. These sites are also included here. In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that makes it difficult, if not impossible, to consistently structure and balance the comparison of a larger number of sites with associated assemblages holding lithic artefacts numbering in the thousands.

## Methodology

As the title of the paper suggest, the data has been analysed from an exploratory perspective, meaning that a wide range of methods has been applied. Only parts of this process can sensibly be reported here. The data and R script used for the analysis is freely available as a research compendium following @marwick , allowing readers to explore the data and the analytical choices made [cf. @Marwick2017]. However, this kind of data dredging does arguably not constitute a good foundation for the pursuit of causal relationships, as this process is most certainly impacted by and selectively reported upon, following my own biases. It can, however, result in the identification of interesting relationships that can be subjected to more rigorous testing using different data and approaches in subsequent studies.

The first part of the analysis conducted involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods. This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995]. Consequently, several artefact categories have here been collapased for the CA. This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading. These have all been combined into the single category "small flint tools". (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption is therefore that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While ultimately intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.

However, for the most part we lack even a most basic understanding of what any individual lithic object has been used for [@dibble2017]. This has major implications that the above-outlined analysis does not take properly into account, rendering it difficult to align any identified pattern with specific behavioural dimensions. For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life. While use-wear analysis could potentially offer a way to identify what artefacts were used for towards the end of their use-life, these kinds of analyses are extremely time-consuming and are therefore typically only conducted on a smaller number of artefacts that have already been selected for analysis based on their shape [e.g. @solheim2018]. Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017]. As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with these inferential limitations in mind [see @clark2017, and below]. The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1979]. An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of tool replacement. Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@binford1979].

However, following not least from the ambiguous definition first put forward by @binford1979, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed, and no single definition has ever been reached [e.g. @bamforth1986; @shott1996; @surovell2009, 9--13]. The continuum between curated and expedient technology has for example been related to dimensions such as land-use and mobility strategies, as well as raw-material quality and availability [e.g. @parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017]. Still, that the distinction can offer a useful analytical framework if clearly operationalised seems more or less agreed upon, and some dimensions of the concept are generally accepted. For example, although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool --- or extending its use-life --- by the repeated rejuvenation and modification of edges [e.g. @bamforth1986; @dibble1995; @shott2005]. Furthermore, one concrete operationalisation of the terms have been forwarded by Barton and collegues [e.g. -@barton1999; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages provide consistent results across a wide range of chronological and cultural context, and is a measure taken to reflect degree of curation. This is in turn mainly to follow from the accumulated nature of the land-use and mobility patterns associated with the sites [@barton2014]. In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades [@clark2017].

These are the main measure employed here. However, as these measures are argued to predominantly be determined by mobility patterns, some additional variables suggested by @Bicho2020 to be related to mobility patterns are also explored. This is done using principal components analysis (PCA), following the continuous operationalisations of these measures [cf. @baxter1994, 100].

A note should also be made on the fact that a couple of variables that are sometimes invoked for the classification of sites in terms of associated mobility pattern are omitted here. This pertains to number of site features such as fireplaces, cooking pits and dwelling structures. Feature count has been omitted as taphonomic loss is likely to have led to a chronological bias in their preservation. Similarly, the number of activity areas, effectively number artefact clusters, however defined, has also been disregarded. This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied. This pertains for example to the impact of bioturbation in the form of three-throws, which can can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jørgensen2017; @darmark2018e].

@@

Figure \@ref(fig:cor) displays the CA using the lithic count data. The general impression from the plots is that a chronological dimension is associated with the patterning in the data. This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data. The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension. Although fewer in number, the sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.

@@

lithics2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
          -askeladden_id) %>% 
  # The reason for not using .keep below is for the option to 
  #retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         blanks_freq = sum(blade, microblade, flake, flake_other, fragment,
                           fragment_other, blade_other,  microblade_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
         -chip_other, -core_other, -core_fragment_other) %>% 
  column_to_rownames("site")

# Repeat but for flint only
flint2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project, 
         -reported_start_bc, -reported_end_bc, -askeladden_id) %>% 
  select(!ends_with("other")) %>% 
  mutate(chip_freq =  sum(chip) / artefact_count,
         core_freq = sum(core) / artefact_count,
         blanks_freq = sum(blade, microblade, flake, fragment) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch) / artefact_count,
         flint_dens = flint_count / excavated_m3) %>%
  select(-blanks_flint, -secondary_flint, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -grinding_slab_stone, -core, 
         -core_fragment, -chip, dating_method) %>% 
  column_to_rownames("site")

@@

```{r cor-plot, fig.cap = "Correlation plot showing transformation of skewed variables for the PCA, and the linear correlation between these", fig.height = 9.5, fig.width = 10}
# Transform the skewed variables - sqrt() is used where there are zero values,
# and seems to provide an adequate result.
lithicstr <- lithics2 %>%
  # Transform skewed variables and reorder
  mutate("Mean date cal. BCE*-1" = mean_date * -1,
          "log10(lithic_dens)" = log10(lithic_dens),
         "log10(secondary_freq)" = log10(secondary_freq),
         "core_freq^(1/3)" = core_freq^(1/3), .keep = "unused") %>% 
  relocate("Mean date cal. BCE*-1",
    "log10(lithic_dens)", "blanks_freq", "log10(secondary_freq)",
    "chip_freq", "core_freq^(1/3)" 
  )
